Bulk Semantic Annotation with a Partially Known Knowledge Base

### 2024-12-05 doc/cicm25-sysdesc/sysdesc.tex
more

\section{Semantic Authoring in a Content Common}
The main practical problem of annotating presentational {\LaTeX} course materials into
semantic documents that carry enough information to support meaningful learning support
services is to add semantic references to technical terms from the domain jargon. Take for
instance a learning object like the paragraph in \cref{fig:lo}, where we want to annotate
the the term ``terminal'' with the (semantic) concept of ``a goal state of a search problem''. The latter can either be defined earlier in the course or in the domain model
(or both).

@@

% In this case the symbol name happens to be the same as its English
% verbalization, otherwise we would have to use
% \lstinline[mathescape]|\sr{$\llangle symbol name\rrangle$}{state space}|.
The annotation is legitimized\ednote{grounded?} by a definition in the domain model module shown in
\cref{fig:state-space}. The \lstinline|\definame| in line 3 introduces the
\textbf{verbalizations} ``goal state'' and ``terminal state''
for the symbol \lstinline|search-problem?goal state|
introduced be \lstinline|\symdecl*| declaration in line 3.

\begin{figure}[ht]\centering
  \fbox{\includegraphics[width=12cm, trim={1cm 0 0 0}, clip]{../img/search-problem.en.pdf}}
\begin{lstlisting}[morekeywords={definame,symdecl},numbers=left]
\begin{smodule}[title=Search Problem]{search-problem}
[... some imports ...]
\symdecl*{goal state}
\begin{sdefinition}
  [...] Certain \sns{state} are [...] \definame[post=s]{goal state} [...]
  (also called \definiendum{goal state}{terminal states}).
\end{sdefinition}
\end{smodule}
\end{lstlisting}
  \caption{Simplified definition of ``goal state'' from the domain model}\label{fig:state-space}
\end{figure}

@@

>>>>

\section{Semantic Authoring in a Content Commons}

Before we discuss semantic authoring in general, let us introduce a concrete example from
the \sTeX/\ALeA world. We will use it to sharpen our intuitions of the problems an issues
involved and to ground the solution we propose.

\subsection{Running Example: Annotating  ``Terminal''}

The main practical problem of annotating presentational {\LaTeX} course materials into
semantic documents that carry enough information to support meaningful learning support
services is to add semantic references to technical terms from the domain jargon. Take for
instance a learning object like the paragraph in \cref{fig:lo}, where we want to annotate
the the term ``terminal'' with the (semantic) concept of ``a goal state of a search
problem''. The latter can either be defined earlier in the course or in the domain model
(or both).

@@

% In this case the symbol name happens to be the same as its English
% verbalization, otherwise we would have to use
% \lstinline[mathescape]|\sr{$\llangle symbol name\rrangle$}{state space}|.
The annotation rerfers to a definition in the domain model module shown in
\cref{fig:state-space}. The \lstinline|\definame| in line 3 introduces the
\textbf{verbalizations} ``goal state'' and ``terminal state'' for the symbol
\lstinline|search-problem?goal state| introduced be \lstinline|\symdecl*| declaration in
line 3.

\begin{figure}[ht]\centering
  \fbox{\includegraphics[width=12cm]{../img/search-problem.en.pdf}}
\begin{lstlisting}[morekeywords={definame,symdecl},numbers=left]
\begin{smodule}[title=Search Problem]{search-problem}
[... some imports ...]
\symdecl*{goal state}
\begin{sdefinition}
  [...] Certain \sns{state} are [...] \definame[post=s]{goal state} [...]
  (also called \definiendum{goal state}{terminal states}).
\end{sdefinition}
\end{smodule}
\end{lstlisting}
  \caption{Simplified definition of ``goal state'' from the domain model}\label{fig:state-space}
\end{figure}

@@

\subsection{Semantic Authoring as a Distinct Problem}

\begin{newpart}{MK: looking at the problem in general and compare it to the informal
    authoring and formal authoring problems e.g. programming in an IDE.}
We contend that semantic authoring is a problem that is distinct from traditional
(informal) authoring and authoring fully formal corpora (e.g. programs or formalizations),
while sharing some of the aspects of either.

In traditional authoring e.g. of a scientific article or a textbook, the domain model -- a
highly structured model of the domain of discourse -- is in the authors' heads, and is
partially verbalized e.g. in the preliminaries section or the main sections of the
respective article. The content is informal -- usually a natural language like english
augmented with technical jargon, formulae, tables, and diagrams -- and designed for
processing via the brain of a colleague or student -- in any case, a human who shares a
canonical part of the domain model with the authors.\ednote{MK: this shared domain model
  could be thought of as the informal content commons; hmmm, but maybe the literature is
  the content commons} Consequently a large part of the authoring problem lies in
predicting the state and extent of the domain model of the reader and creating text that
bridges the gap between the reader's domain model and the payload content of the article
-- the knowledge the article or textbook intends to impart -- ideally using the technical
vocabulary the reader is familiar with to reduce this gap. Tool support for traditional
authoring is minimal and usually restricted to spell/grammar-checkers and online thesauri
if we disregard LLM writing/formulation support for now.

In fully formal authoring e.g. for programming, the content commons is given formally in
the form of the libraries of the respective programming language, correspondingly the
authored content is also formal -- a program -- and intended for processing by a computer,
for which the formality is a prerequisite.

\ednote{talk about formalization in logic -- e.g. as theorem prover libraries as well}
\end{newpart}
